`flippant` -- An `R` package for the automated analysis of dithionite scramblase

### 2015-09-21 inst/manuscript/manuscript.rmd
clarify knitr options;
More edits/typo fixing;

knitr::opts_chunk$set(fig.width=6.78, fig.height=6.78, warning=FALSE, message=FALSE)
```

@@

To interrogate this functional class of proteins, we have developed a biochemical 
assay suited to characterize scramblase activity of purified proteins and protein 
extracts based on the reconstitution of scramblases or candidate proteins into 
preformed synthetic liposomes that contain trace amounts of nitrobenzoxadiazole 
(NBD) labeled fluorescent lipids [@chang_chemical_2004].
Upon addition of membrane-impermeable reducing agents such as dithionite to 
these liposomes, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence disappears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the 
assay has also been used to characterize the stoichiometry of scramblase 
homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the 
package ...**

@@

Based on @menon_opsin_2011 and @goren_constitutive_2014, data is processed as 
follows. Input is format checked and defaults are injected for facultative 
parameters/columns as appropriate. Fluorescence spectra are parsed. This includes 
algorithmic determination of when dithionite was added to the sample using peak
detection by continuous wavelet transformation as implemented in the 
`wmtsa` `R` package [@percival_wavelet_2000;@constantine_wmtsa:_2013]. 
Acquisition time is aligned for all spectra in the data set such that the zero 
time point henceforth corresponds to the time of addition. 
Pre-dithionite-addition baseline fluorescence $F_{\text{baseline}}$ is 
determined for each spectrum by averaging (median) over the ten values preceding
dithionite addition. Post-dithionite-addition minimum fluorescence 
$F_{\text{min}}$ is analogously calculated from the last ten data points after 
a default 400 s or the time point of measurement supplied by the user in the 
parameter spreadsheet. $F_{\text{min}}$ is volume-corrected based on supplied 
reaction volumes with and without dithionite and for each spectrum/data point a 
fluorescence reduction is calculated as follows:
$$
y_{\text{raw}} = 1 - \frac{F_{\text{min}}}{F_{\text{baseline}}}
$$
A relative fluorescence reduction $y$ is calculated scaled to a (required) liposomes-only/no-protein control and the probability $P_{\geq 1}$ for a 
liposome to contain at least one scramblase molecules is calculated using
$$
P_{\geq 1} = \frac{(y - y_0)}{(y_{\text{max}} - y_0)}
$$
where $y$ is the relative fluorescence reduction for the proteoliposome 
measurement and $y_0$ for a measurement of protein-free liposomes. 
$y_{\text{max}}$ refers to either the maximal $y$ in the experiment or is 
derived from a monoexponential fit to the data (default) as a precaution for the 
case where the protein/phospholipid titration did not reach the plateau of the 
saturation curve. The behavior is adjustable using the `scaleTo` parameter of 
the corresponding functions (see below). Next a monoexponential curve is fitted
to the relationship of $P_{\geq 1}$ to the protein to phospholipid ratio (PPR; 
(mg/mmol)), providing the scramblase-characterizing fit factor $\alpha$.

@@

#Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, of which the 
majority of cases is linked to mutant forms of the G-protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by our recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved when tested using the dithionite scramblase assay
among other techniques. However, the same set of experiments pointed at a defect 
in rhodopsin dimerization prior to insertion into liposomes, generating the new 
hypothesis that the mutations may cause disease by interfering with the highly 
ordered quarternary structure of the molecule in the retina 
[@ploier_perturbed_submitted].

>>>>

knitr::opts_chunk$set(
  fig.width=6.78, 
  fig.height=6.78, 
  warning=FALSE, 
  message=FALSE,
  dev = 'pdf')
```

@@

To interrogate scramblases as a functional class of proteins, we have developed 
a biochemical assay suited to characterize scramblase activity of purified 
proteins and protein extracts based on the reconstitution of scramblases or 
candidate proteins into preformed synthetic liposomes that contain trace amounts
of nitrobenzoxadiazole (NBD) labeled fluorescent lipids [@chang_chemical_2004].
Upon addition of membrane-impermeable reducing agents such as dithionite to 
these liposomes, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence disappears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the 
assay has also been used to characterize the stoichiometry of scramblase 
homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the 
package ...**

@@

Based on @menon_opsin_2011 and @goren_constitutive_2014, data is processed as 
follows. Input is format checked and defaults are injected for facultative 
parameters/columns as appropriate. Fluorescence spectra are parsed. This includes 
algorithmic determination of when dithionite was added to the sample using peak
detection by continuous wavelet transformation as implemented in the 
`wmtsa` `R` package [@percival_wavelet_2000;@constantine_wmtsa:_2013]. 
Acquisition time is aligned for all spectra in the data set such that the zero 
time point henceforth corresponds to the time of addition. 
Pre-dithionite-addition baseline fluorescence $F_{\text{baseline}}$ is 
determined for each spectrum by averaging (median) over the ten values preceding
dithionite addition. Post-dithionite-addition minimum fluorescence 
$F_{\text{min}}$ is analogously calculated from the last ten data points before 
(and including) a default 400 s or the time point of measurement supplied by the
user in the parameter spreadsheet. $F_{\text{min}}$ is volume-corrected based on
supplied reaction volumes with and without dithionite and for each spectrum/data
point a fluorescence reduction is calculated as follows:
$$
y_{\text{raw}} = 1 - \frac{F_{\text{min}}}{F_{\text{baseline}}}
$$
A relative fluorescence reduction $y$ is calculated scaled to a (required) liposomes-only/no-protein control and the probability $P_{\geq 1}$ for a 
liposome to contain at least one scramblase molecules is calculated using
$$
P_{\geq 1} = \frac{(y - y_0)}{(y_{\text{max}} - y_0)}
$$
where $y$ is the relative fluorescence reduction for the proteoliposome 
measurement and $y_0$ for a measurement of protein-free liposomes. 
$y_{\text{max}}$ refers to either the maximal $y$ in the experiment or is 
derived from a monoexponential fit to the data (default) as a precaution for the 
case where the protein/phospholipid titration did not reach the plateau of the 
saturation curve. The behavior is adjustable using the `scaleTo` parameter of 
the corresponding functions (see below). Next a monoexponential curve is fitted
to the relationship of $P_{\geq 1}$ to the protein to phospholipid ratio (PPR; 
(mg/mmol)), providing the scramblase-characterizing fit factor $\alpha$.

@@

#Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G-protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by our recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved when tested using the dithionite scramblase assay
among other techniques. However, the same set of experiments pointed at a defect 
in rhodopsin dimerization prior to insertion into liposomes, generating the new 
hypothesis that the mutations may cause disease by interfering with the highly 
ordered quarternary structure of the molecule in the retina 
[@ploier_perturbed_submitted].
